---
title: "Exercise 03"
author: "Hubert Rehrauer"
date: "04 10 2021"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(tibble)
library(edgeR)
library(matrixStats)
library(reshape2)
library(stats)
library(pheatmap)
library(ggfortify)
library(RColorBrewer)
knitr::opts_chunk$set(echo = TRUE)
```


## Exploratory Data Analysis

Do an exploratory data analysis of a matrix of expression values. The data consists of expression values for samples that were treated with DMSO and TSA. The samples were measured using three technologies: bulk, IFC96, IFC800. See the two RDS files `counts.RDS` and `phenodata.RDS`


## Data Import
```{r data import, echo=FALSE}
x = readRDS("/Users/serenagalli/Dropbox/Computational_Science/STA426/week03-04oct2021/exercise03/counts.RDS")
anno = readRDS("/Users/serenagalli/Dropbox/Computational_Science/STA426/week03-04oct2021/exercise03/phenodata.RDS")
```


## Compute and visualize basic statistics
How many replicates are there for each combination of factor levels?
```{r basic statistics, echo=TRUE}
table(anno)
```
How many genes have an expression value above 0 in each sample?
```{r basic statistics 2, echo = TRUE}
colSums(x != 0)
```


## Normalize the data
Scale the columns so that the total sum of all columns are identical
```{r scaling}
scaled_x <- cpm(x)
```


## Transform the data to log-scale
Use the function `log1p` to transform the data to log-scale
```{r normalization}
transformed_x <- log1p(scaled_x)
```


## Visualize the distribution of the expression values
Use violin plots and boxplots to visualize the distribution of the expression values
Group and color by experimental factors.

Unfortunately, my deficit in basic R skills is bigger than I thought and I managed neither to define the 6 experimental groups nor to bring the dataframe in a form that can be used for plotting the graphs. For the plots, I wrote down a code, where x stands for thedataframe in the correct form.

This is what I tried: 

rownames(anno) <- colnames(x)
df_merged <- merge(x, t(anno))
ggplot(x, aes(fill= sample, x= experimental_factor, y= expression, color= Technology)) +
  geom_violin() +
  geom_boxplot(width=0.1,color="black",position = position_dodge(width =1))


## Most variable genes
Identify the 500 most variable genes (with largest variance across samples) and continue working with those

```{r most variable features, echo=TRUE}
rv <- rowVars(transformed_x)
ind_top500 <- order(rv, decreasing=TRUE)[seq_len(min(500, length(rv)))]
top_500 <- transformed_x[ind_top500,]
```


## Sample correlations
Compute and visualize the sample-to-sample correlations
```{r correlation matrix, echo = TRUE}
cor_matrix <- round(cor(top_500), 2)
melted_cor_matrix <- melt(cor_matrix)
ggplot(data = melted_cor_matrix, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() + theme(axis.title.y=element_blank(),
        axis.text.y=element_blank())

```




## Clustering
Compute and visualize a hierarchical clustering of the samples, use package `hclust`

```{r clustering}
gene_dist <- dist(t(top_500))
clustered <- hclust(gene_dist, method = "complete", members = NULL)
plot(clustered, labels = FALSE)
```



## Heatmap
Use the package `pheatmap` to generate a heatmap of the expression data.
```{r heatmap}
pheatmap(top_500,
        colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100),
        cluster_rows = T,
        show_rownames = F,
        scale = "row",
        )
```




## PCA

In the exercise session, we saw a potential case where the normal PCA can be misleading. 

* Generate the data on the left, plot it as a scatter plot and color the points based on their class label. 
* Do the PCA, plot the variance explained by the principal components. Select $k$ such that you explain $80\%$ of the variance in your data. 
* Explain why this problem happens. 
* Try to solve the problem with your desired method and do the PCA, and plot the variance explained by your PCs. Select $k$ in order to explain $80\%$ of the variance in your data. Is the $k$ different in comparison to the previous case? 
 
I didn't figure out the difference between computing a regular PCA and a non-regular PCA. Since I got more than 2 dimension, I guess that I computed a non-regular PCA. 

```{r PCA,  echo=TRUE}
top_500.pca <- prcomp(top_500, center = TRUE,scale. = TRUE)
summary(top_500.pca) # cumulative proportion of variance > 80% for PC 35, i.e. k = 35
cumpro <- cumsum(top_500.pca$sdev^2/sum(top_500.pca$sdev^2))
plot(cumpro, xlab = "PC #", ylab ="Amount of explained variance", main = "Cumulative variance plot")
autoplot(top_500.pca)
```





